### 2012-08-25 paper.tex
suggested changes to first draft

Various papers have been written on tennis-related injuries from a physiological perspective.  In particular, \cite{demands} attempted to quantify the demands in professional male tennis by analysing the number and type of strokes played per game for 22 players from three Grand Slams.  They found that the serve was the predominant stroke played in service games (up to 60\%) whereas topspin forehands and backhands were more frequent when receiving.  The 2003 US Open winner hit over 1000 serves during his seven matches at the tournament.  More strokes are played at the French Open than Wimbledon due to the relative speeds of the clay and grass court surfaces leading to longer rallies at Roland Garros.  Importantly, Johnson and McHugh discuss the strain that playing a point inflicts on the body.  They report that over 50\% of world-class tennis players experience shoulder discomfort during their career and 80\% of these cases stem from overuse.  Stroke production in tennis involves generating repetitive forces and motions that are of high intensity and short duration.  For example, the serve is the most strenuous stroke on the upper extremity with internal rotation velocities of the humerus reaching 2420\% for elite players during the acceleration phase.  This, coupled with the relentless demands the ATP (Association of Tennis Professionals) and WTA (Women's Tennis Association) tours place on players, suggest that it is no surprise injuries are an issue in the world of professional tennis.

@@

where $r_n$ is the given player's risk of retirement on point $n$ of the match, $0 \leq \rho \leq 1$, and $X$ and $Y$ are random variables.  Do not confuse $r_n$ with what we are hoping to eventually calculate which is $R_n$, the risk of retiring at some point during the remainder of the match from point $n$.  We say that players start a match with zero probability of retiring ($r_0 = 0$), else they would choose to not participate and allow their opponent a walkover.  This is not strictly true since players sometimes play matches despite possessing niggling injuries but it is a reasonable assumption.  When setting $r_{n+1}$, we make sure to take the minimum of the calculated $r_{n+1}$ and 1, since retirement risk is a probability and cannot be greater than 1.  The Bernoulli random variable, $X$, models the chance of an injury occurring on a given point.  The random variable, $Y$, models the magnitude of an injury should it occur and is truncated exponentially distributed with upper bound 1, i.e. minor bumps and bruises are more common than serious, match-ending injuries.  The decay parameter, $\rho$, models the idea that players recover from injuries as matches progress.  For the purposes of avoiding too complex a model, we make the simplifying assumption that $\rho$ and the distributions of $X$ and $Y$ are the same for both players.

@@

A given point in a simulated match, has four possible outcomes.  Either player can win the point or either player can retire from the match.  To resolve a point, we generate a random number between 0 and 1 and observe which of the bins (as shown in Figure \ref{bins}), the number falls in.   The match ends if it falls in one of the retirement bins (which is why $r_A$ and $r_B$ are usually 0 else you will rarely be able to get through even a single match without retiring), otherwise one of the players wins the point and the match continues (unless it was match point).

\begin{figure}[H]\LARGE
	\begin{center}
		\begin{tabular}{|C{3.5cm}|C{2.5cm}|C{0.7cm}|C{0.7cm}|}
			\multicolumn{1}{l}{\small{0}} & \multicolumn{1}{r}{} & \multicolumn{1}{r}{} & \multicolumn{1}{r}{\small{1}} \\
			\hline
			$p$ & $q$ & \Large{$r_A$} & \Large{$r_B$}\\ \hline
		\end{tabular}
	\end{center}
	\caption{The four possible outcomes of a point in our modified tennis match simulator}
	\label{bins}
\end{figure}

To imitate the Betfair Set Betting market (or equivalently, a Match Odds market using a Paddy Power-style \textit{match-completed} payout policy), we can use O'Malley's tennis formulae which ignore retirement risk (the \textit{No Risk} column in Figure \ref{outcomes}).  We use our tennis match simulator to imitate a Match Odds market as it might behave using the different retirement betting payout policies by calculating the remaining probabilities in the right-most three columns.

\begin{figure}[H]
	\begin{center}
	\renewcommand{\arraystretch}{1.8}
		\begin{tabular}{|C{1.2cm}|C{1.4cm}|C{2.4cm}|C{2.4cm}|C{2.4cm}|}
			\hline
			\textbf{Player} & \textbf{No Risk} & \textbf{Normal Win With Risk} & \textbf{Retirement in 1st set} & \textbf{Retirement after 1st Set} \\ \hline
			$A$ & $W_A$ & $W'_A$ & $R'_A$ & $R^{''}_A$ \\ \hline
			$B$ & $W_B$ & $W'_B$ & $R'_B$ & $R^{''}_B$ \\ \hline
		\end{tabular}
	\renewcommand{\arraystretch}{1}
	\end{center}
	\caption{Probabilities that can be closely approximated by our modified tennis match simulator}
	\label{outcomes}
\end{figure}

@@

where $R_A = R'_A + R^{''}_A$ and $R_B = R'_B + R^{''}_B$.  This is essentially a re-calculation of the \textit{Normal Win With Risk} column.

@@

We have introduced three unknowns into our simulator; $c$ is the success parameter of the Bernoulli distribution (per point injury probability), $\lambda$ is the injury magnitude truncated exponential distribution rate parameter, and $\rho$ is the injury recovery factor.

Figure \ref{variabletable} displays a table describing the majority of the variables we have introduced thus far.  An appropriate parameterisation of the model would be values for the unknown input variables, $c$, $\lambda$, and $\rho$, that generate $R_A \approx R_B \approx 1.95\%$ when input into our modified simulator at the start of a 5-set match with $P_A = P_B = 0.6$.  We believe this is justified as it corresponds to the 3.9\% chance of such a match ending in retirement as seen in men's Grand Slam singles matches between 2000 and 2009, as well as a common point-winning probability on serve of a top-level professional tennis player.  The goal now is to \textit{fit} the model to accurately reflect real-world events.

It is possible to use a multivariate direct-search optimisation algorithm such as the Nelder-Mead simplex method (an algorithm primarily designed for statistical parameter estimation problems such as ours) to approximate a set of values for these parameters.  The method uses the idea of a \textit{simplex}, a polytope of $N+1$ vertices given $N$ dimensions, i.e. a line segment on a line, a triangle on a plane, a tetrahedra in 3D space, etc.  The user defines an initial non-degenerate simplex and an \textit{objective function} that the algorithm attempts to minimise. Specifically, we used a constrained version of the Nelder-Mead algorithm in order to ensure logical bounds $0 < c, \rho < 1$ and $\lambda > 0$.

@@

Vitally important to the success of the Nelder-Mead method is the choice of objective function to minimise.  In our case, our modified simulator is essentially the function, but we must still define what it means to minimise it.  This is the main way we mould the algorithm to solve our problem.

\begin{center}
	$\left|W_A - (W'_A + R_A)\right| + \left|W_B - (W'_B + R_B)\right|$
\end{center}

After the algorithm has completed, the first term ensures that we have recreated the correct gap in the odds for Player A with the second term being the same idea for Player B.  So in this case, at the minimum, we would have:

@@

We executed the Nelder-Mead method a number of times under the same conditions and took the means of the approximations found for our three parameters each time as our chosen values.  Remember, $c$ is the per point injury probability, $\lambda$ is the point-level retirement risk magnitude exponential distribution rate parameter, and $\rho$ is the injury recovery factor.

\begin{itemize}
	\item $c = 0.000115$ (implying an injury approximately every 8500 points)
	\item $\lambda = 10.0$ (implying injuries cause a point-level retirement risk of 0.1 on average when they occur)
	\item $\rho = 0.95$ (point-level retirement risk is scaled by 0.95 each point)
\end{itemize}

@@

In order to find out what our model is capable of, we try it out in a totally artificial environment where we control all the variables.  We ran a single simulated match many times with our chosen set of parameters.  We assume that $P_A$ and $P_B$ remain unchanged throughout the match.  We are looking for 'ideal' scenarios, e.g. where Player A receives an injury during the match but does not retire and still goes on to achieve victory.  When we generate such a match, we recorded into a CSV file the score, the server, and the point-level retirement risks, $r_A$ and $r_B$, at each point during the match.  We then read this information back in, calculating match-winning probabilities at each point in the match.

@@

Up until this point, we have assumed that we will be able to manually input the point-winning probabilities on serve of both players in the match.  We note the findings of \cite{dominance} that the average point-winning probability on serve for a top-level professional tennis player, $\gamma$, was 0.645 for men and 0.560 for women.  We also know, according to \cite{marek}, that the important thing in determining the winner of a match is the \textit{difference}, $\delta$, between the point-winning probabilities of each player and not the absolute values.  Consequently, if we can find what $\delta$ \textit{should} be, we can choose appropriate values for $P_A$ and $P_B$ by using the constraint that their average equals $\gamma$.  We must find both; we cannot, for example, fix $P_A$ to $\gamma$ and linearly search for $P_B$, since the model would cease to be symmetric for both players.  We know the current implied match-winning probability of the Set Betting (no retirement risk) market for both players and so we can do a simple binary search to find a value for $\delta$ (and therefore values for $P_A$ and $P_B$) for which O'Malley's tennis formulae generate the same results as the \textit{no risk} market.  Marek also states that for one to be able to express match-winning probability as a function of $\delta$, one must constrain $-0.1 \leq \delta \leq 0.1$.  Since the market's opinion of each player's point-winning probability changes throughout the match, we recalculate $P_A$ and $P_B$ on every point.  Note that this does not invalidate the assumption that points are iid as no dependency between points is introduced.

@@

As the match progresses and Nadal falls further behind, we see spikes in his retirement risk despite there being no large gaps between the Betfair markets.  This happens when the Match Odds market gives a player little chance of winning the match.  We illustrate how this occurs with an example.  Say that on a particular point, the Set Betting market tells us that the implied probability of Nadal winning the match is 0.5.  This means that we can assign $P_A = P_B = 0.645$.  We also happen to know that the difference between the Betfair Set Betting and Match Odds for Nadal is 0.4.  We now look for a value of $r_A$ (retirement risk of Nadal on this point) such that $W_A^{''}$ (probability of winning the match normally with retirement risk after the first set only) is only 0.1.  Since we have $P_A = P_B$, we have $W_B^{''} = 0.1$ and so the sum of the probabilities of either player winning the match normally is only 0.2.  Making the assumption that $r_n^B$ is negligible, we must have that $R_A \approx 0.8$ (retirement risk of Nadal in the match).  This can happen in any situation where $W^{''}$ is small and there is a risk of retirement, like towards the end of this quarter final.

@@

We have created a new model for tennis in the form of a tennis match simulator which takes into account the retirement risk of players by incorporating extra parameters approximated using real-world averages and betting odds data as well as additional match outcomes compared to standard tennis models.  This is the world's first attempt to create such a model.  Testing the model in a totally artificial environment, we found it was able to mimic the expected patterns of in-play betting markets with different retirement payout policies for matches with varying injury scenarios.

@@

>>>>

Various papers have been written on tennis-related injuries from a physiological perspective.  In particular, \cite{demands} attempted to quantify the demands in professional male tennis by analysing the number and type of strokes played per game for 22 players from three Grand Slams.  They found that the serve was the predominant stroke played in service games (up to 60\%) whereas topspin forehands and backhands were more frequent when receiving.  The 2003 US Open winner hit over 1000 serves during his seven matches at the tournament.  More strokes are played at the French Open than Wimbledon due to the relative speeds of the clay and grass court surfaces leading to longer rallies at Roland Garros.  Importantly, Johnson and McHugh discuss the strain that playing a point inflicts on the body.  They report that over 50\% of world-class tennis players experience shoulder discomfort during their career and 80\% of these cases stem from overuse.  Stroke production in tennis involves generating repetitive forces and motions that are of high intensity and short duration.  For example, the serve is the most strenuous stroke on the upper extremity with internal rotation velocities of the humerus reaching 2420 degrees per second for elite players during the acceleration phase.  This, coupled with the relentless demands the ATP (Association of Tennis Professionals) and WTA (Women's Tennis Association) tours place on players, suggest that it is no surprise injuries are an issue in the world of professional tennis.

@@

where $r_n$ is the given player's risk of retirement on point $n$ of the match, $0 \leq \rho \leq 1$, and $X$ and $Y$ are random variables.  Do not confuse $r_n$ with what we are hoping to eventually calculate which is $R_n$, the risk of retiring at some point during the remainder of the match from point $n$.  We say that players start a match with zero probability of retiring ($r_0 = 0$), although in principle it is possible players might start a match with a niggling injury, in which case the magnitude could be inferred from observations of two or more match markets with differing retirement policies.  When setting $r_{n+1}$, we make sure to take the minimum of the calculated $r_{n+1}$ and 1, since retirement risk is a probability and cannot be greater than 1.  The Bernoulli random variable, $X$, with success parameter $c$, models the chance of an injury occurring on a given point.  The random variable, $Y$, models the magnitude of an injury should it occur and is truncated exponentially distributed with rate parameter $\lambda$ and upper bound 1, i.e. minor bumps and bruises are more common than serious, match-ending injuries.  The decay parameter, $\rho$, models the idea that players recover from injuries as matches progress.  For the purposes of avoiding too complex a model, we make the simplifying assumption that $\rho$ and the distributions of $X$ and $Y$ are the same for both players.

@@

A given point in a simulated match has four possible outcomes.  Either player can win the point or either player can retire from the match.  To resolve a point, we generate a random number between 0 and 1 and observe which of the bins (as shown in Figure \ref{bins}), the number falls in.   The match ends if it falls in one of the retirement bins (which is why $r_A$ and $r_B$ are usually 0 else you will rarely be able to get through even a single match without retiring), otherwise one of the players wins the point and the match continues (unless it was match point).

\begin{figure}[H]\Large
	\begin{center}
		\begin{tabular}{|C{3.5cm}|C{2.5cm}|C{0.7cm}|C{0.7cm}|}
			\multicolumn{1}{l}{\small{0}} & \multicolumn{1}{r}{} & \multicolumn{1}{r}{} & \multicolumn{1}{r}{\small{1}} \\
			\hline
			$p$ & $q$ & $r_A$ & $r_B$ \\ \hline
		\end{tabular}
	\end{center}
	\caption{The four possible outcomes of a point in our modified tennis match simulator}
	\label{bins}
\end{figure}

To imitate the Betfair Set Betting market (or equivalently, a Match Odds market using a Paddy Power-style \textit{match-completed} payout policy), we can use O'Malley's tennis formulae which ignore retirement risk (the \textit{No Retirement Risk} column in Figure \ref{outcomes}).  We use our tennis match simulator to imitate a Match Odds market as it might behave using the different retirement betting payout policies by calculating the remaining probabilities in the right-most three columns.

\begin{figure}[H]
	\begin{center}
	\renewcommand{\arraystretch}{1.8}
		\begin{tabular}{|C{1.2cm}|C{1.4cm}|C{2.8cm}|C{2.4cm}|C{2.4cm}|}
			\hline
			\textbf{Player} & \textbf{No Ret. Risk} & \textbf{Normal Win With Ret. Risk} & \textbf{Retirement in 1st set} & \textbf{Retirement after 1st Set} \\ \hline
			$A$ & $W_A$ & $W'_A$ & $R'_A$ & $R^{''}_A$ \\ \hline
			$B$ & $W_B$ & $W'_B$ & $R'_B$ & $R^{''}_B$ \\ \hline
		\end{tabular}
	\renewcommand{\arraystretch}{1}
	\end{center}
	\caption{Probabilities that can be closely approximated by our modified tennis match simulator}
	\label{outcomes}
\end{figure}

@@

where $R_A = R'_A + R^{''}_A$ and $R_B = R'_B + R^{''}_B$.  This is essentially a re-calculation of the \textit{Normal Win With Retirement Risk} column.

@@

We have introduced three unknowns into our simulator; $c$ is the success parameter of the Bernoulli distribution (per point injury probability), $X$, $\lambda$ is the rate parameter of the injury magnitude truncated exponential distribution, $Y$, and $\rho$ is the injury recovery factor.

Figure \ref{variabletable} displays a table describing the majority of the variables we have introduced thus far.  An appropriate parameterisation of the model would be values for the unknown input variables, $c$, $\lambda$, and $\rho$, that generate $R_A \approx R_B \approx 1.95\%$ when input into our modified simulator at the start of a 5-set match.  We believe these match-level retirement risks are justified as it corresponds to the 3.9\% chance of such a match ending in retirement as seen in men's Grand Slam singles matches between 2000 and 2009.  Given \cite{dominance} find that the average point-winning probability on serve for a top-level professional tennis player is 0.645 for men and 0.560 for women, we choose 0.6 as a reasonable value for $P_A$ and $P_B$ which denotes an even match.  The goal now is to \textit{fit} the model to accurately reflect real-world events.

It is possible to use a multivariate direct-search optimisation algorithm such as the Nelder-Mead simplex method (an algorithm primarily designed for statistical parameter estimation problems such as ours) to approximate a set of values for these parameters.  Specifically, we used a constrained version of the Nelder-Mead algorithm in order to ensure logical bounds $0 < c, \rho < 1$ and $\lambda > 0$.

@@

Vitally important to the success of the Nelder-Mead method is the choice of objective function to minimise.  In our case, our modified simulator is essentially the function, but we must still define what it means to minimise it.  This is the main way we mould the algorithm to solve our problem.  We want the difference between the probability Player A wins the match normally given no risk of retirement and the probability Player A wins the match normally with risk of retirement to be $0.0195$, which is the probability Player A wins the match via Player B retiring (and similarly for Player B).  Consequently, we aim to minimise the below expression:

\begin{center}
	$\left|W_A - (W'_A + R_B)\right| + \left|W_B - (W'_B + R_A)\right|$
\end{center}

Specifically, at the minimum and since the match is even, we would have:

@@

We executed the Nelder-Mead method a number of times under the same conditions and took the means of the approximations found for our three parameters each time as our chosen values.  Remember, $c$ is the per point injury probability, $\lambda$ is the point-level retirement risk magnitude exponential distribution rate parameter, and $\rho$ is the injury recovery factor.  We find:

\begin{itemize}
	\item $c = 0.000115$ (implying an injury approximately every 8500 points)
	\item $\lambda = 10.0$ (implying injuries cause a point-level retirement risk of 0.1 on average when they occur)
	\item $\rho = 0.95$ (point-level retirement risk is multiplied by 0.95 on each point)
\end{itemize}

@@

In order to assess the accuracy of our model, we try it out in a totally artificial environment where we control all the variables.  We ran a single simulated match many times with our chosen set of parameters.  We assume that $P_A$ and $P_B$ remain unchanged throughout the match.  We are looking for 'ideal' scenarios, e.g. where Player A receives an injury during the match but does not retire and still goes on to achieve victory.  When we generate such a match, we recorded into a CSV file the score, the server, and the point-level retirement risks, $r_A$ and $r_B$, at each point during the match.  We then read this information back in, calculating match-winning probabilities at each point in the match.

@@

Up until this point, we have assumed that we will be able to manually input the point-winning probabilities on serve of both players in the match.  Due to \cite{dominance}, we know that the average point-winning probability on serve for a top-level professional tennis player, $\gamma$, is 0.645 for men and 0.560 for women.  We also know, according to \cite{marek}, that the important thing in determining the winner of a match is the \textit{difference}, $\delta$, between the point-winning probabilities of each player and not the absolute values.  Consequently, if we can find what $\delta$ \textit{should} be, we can choose appropriate values for $P_A$ and $P_B$ by using the constraint that their average equals $\gamma$.  We must find both; we cannot, for example, fix $P_A$ to $\gamma$ and linearly search for $P_B$, since the model would cease to be symmetric for both players.  We know the current implied match-winning probability of the Set Betting (no retirement risk) market for both players and so we can do a simple binary search to find a value for $\delta$ (and therefore values for $P_A$ and $P_B$) for which O'Malley's tennis formulae generate the same results as the \textit{no risk} market.  Marek also states that for one to be able to express match-winning probability as a function of $\delta$, one must constrain $-0.1 \leq \delta \leq 0.1$.  Since the market's opinion of each player's point-winning probability changes throughout the match, we recalculate $P_A$ and $P_B$ on every point.  Note that this does not invalidate the assumption that points are iid as no dependency between points is introduced.

@@

As the match progresses and Nadal falls further behind, we see spikes in his retirement risk despite there being no large gaps between the Betfair markets.  This happens when the Match Odds market gives a player little chance of winning the match.  We illustrate how this occurs with an example.  Say that on a particular point, the Set Betting market tells us that the implied probability of Nadal winning the match is 0.5.  This means that we can assign $P_A = P_B = 0.645$.  We also happen to know that the difference between the Betfair Set Betting and Match Odds for Nadal is 0.4.  We now look for a value of $r_A$ (retirement risk of Nadal on this point) such that $W_A^{''}$ (probability of winning the match normally with retirement risk after the first set only) is only 0.1.  Since we have $P_A = P_B$ and therefore both players have an equal chance of winning the match if it does not end in retirement, we have $W_B^{''} = 0.1$ and so the sum of the probabilities of either player winning the match normally is only 0.2.  Making the assumption that $r_n^B$ is negligible, we must have that $R_A \approx 0.8$ (retirement risk of Nadal in the match).  This can happen in any situation where $W^{''}$ is small and there is a risk of retirement, like towards the end of this quarter final.

@@

We have created a new model for retirement risk in tennis matches and implemented it into a tennis match simulator by incorporating extra parameters approximated using real-world averages and betting odds data as well as additional match outcomes compared to standard tennis models.  This is the world's first attempt to create such a model.  Testing the model in a totally artificial environment, we found it was able to mimic the expected patterns of in-play betting markets with different retirement payout policies for matches with varying injury scenarios.

@@

We note that a potentially useful application of our model is in the real-time estimates of player retirement risk during professional tennis matches.  Such information could be displayed on-screen for the viewer during a live broadcast or used as a commentary aid.
